Extracting sunbeams from cucumbers: Reproducibility as a necessity for the future of archaeological science

### 2025-01-13 paper.qmd
where did my text go?

To objectively quantify the relative hardness or softness of archaeology, as an evaluation of its status as a science, and the place of this journal in context of other archaeology journals, I take a bibliometric approach. This approach is based on @fanelliBibliometricEvidenceHierarchy2013, who examined the hardness and softness of 12 disciplines using scholarly publication parameters. @fanelliBibliometricEvidenceHierarchy2013 found a spectrum of statistically significant variation in bibliometric variables from the physical to the social sciences, with papers tending at the softer end of the spectrum tending to have fewer co-authors, use less substantive titles, have longer texts, cite older literature, and have a higher diversity of sources. In @fanelliBibliometricEvidenceHierarchy2013's analysis harder sciences include Space Science, Physics, Chemistry, less hard sciences include social sciences include Psychiatry/Psychology, Economics, Business and General Social Sciences, and the Humanities define the soft end of the spectrum. Following @fanelliBibliometricEvidenceHierarchy2013, I quantify the number of authors, length of article, relative title length, age of references, and diversity of references for a large sample of peer-reviewed journal articles.

These parameters are useful because of how they signify consensus in a research community. A larger number of authors on a paper reflects collaboration of people working together on a common goal. Collaborators have specialized roles, each of whom has the ability to study a part of the problem with high accuracy and detail, with harder fields having larger groups of collaborators[@zuckerman1972age]. Reflecting this collaboration group size, harder disciplines tend to have higher average numbers of authors on papers. Article length has an inverse correlation with field hardness. In low-consensus, or softer, fields, papers must be longer to present justification, nuance and contextualization of results. The number of substantive and informative words in an article's title tends to be positively correlated with article length in harder disciplines [@yitzhakiRelationTitleLength2002; @yitzhakiVariationInformativityTitles1997], reflecting a focus on empiricism and efficiency that is characteristic of high-consensus disciplines. While @yitzhakiRelationTitleLength2002 removed stop-words (e.g. prepositions, articles, conjunctions, etc.) to calculate article length, in order to generate results for comparison with @fanelliBibliometricEvidenceHierarchy2013 I follow his method of dividing the total word count of the article title by the total number of pages of the article to compute relative title length.

@@

I sampled `r n_articles` papers published during `r year_min`-`r year_max`. This sample was collected from Clarivate's Web of Science database by first selecting the Web of Science category 'Archaeology' and the Document type 'article' (n = 28,871). To focus on journals of broad relevance to most archaeologists, and that are representative of substantial communities of practice, I then filtered the results to keep only articles published in the top-ranking 25 journals according to their h-indices as reported by Clarivate's Journal Citation Indicator. Finally I excluded journals with less than 100 articles in the database.

@@

# Price's index - recency of references ------------------
library(stringr)

@@

plot_grid(boxlplot_n_authors, 
          boxlplot_rel_title_length,
          boxlplot_n_pages,
          boxlplot_price_index, 
          boxlplot_shannon_index,
          nrow = 2)
```

@fig-compare-other-fields shows the distribution of bibliometric variables for archaeology in context of data from other fields presented by @fanelliBibliometricEvidenceHierarchy2013. The most striking indicator of archaeology as a hard science is the number of authors, where it is between the social sciences and physics. Archaeology is a close fit with the social sciences in relative title length. It is between the the social sciences and humanities in recency of references and diversity of references. The clearest indicator of archaeology as a soft science is article length where it is like the humanities. Overall archaeology does not sit squarely at either end of the hard-soft spectrum. It is generally not a harder science than the social sciences, with the exception of collaborator group sizes.

@@

# compute beta estimates so we can colour lines to indicate more or
# less scientific
library(broom)
over_time_long_models <- 
  over_time_long %>% 
  group_nest(variable) %>% 
  mutate(model = map(data, ~tidy(lm(value ~ year, data = .)))) %>% 
  unnest(model) %>% 
  filter(term == 'year') %>% 
  mutate(becoming_more_scientific = case_when(
    variable == "N. of authors (ln)"         & estimate > 0 ~ "TRUE",
    variable == "N. of pages (ln)"           & estimate < 0 ~ "TRUE",
    variable == "N. of refs (sqrt)"          & estimate < 0 ~ "TRUE",
    variable == "Recency of references"      & estimate > 0 ~ "TRUE",
    variable == "Relative title length (ln)" & estimate > 0 ~ "TRUE",
    variable == "Diversity of references"    & estimate < 0 ~ "TRUE",
    TRUE ~ "FALSE"
  ))

# join with data
over_time_long_colour <- 
  over_time_long %>% 
  left_join(over_time_long_models)
```

```{r}
#| label: fig-change-over-time
#| fig-cap: Distribution of article characteristics for archaeology articles over time. Individual articles are shown by data points. The Ordinary Least Squares method was used to fit the straight lines summarising the relationships between the variables and the time series.     
ggplot(over_time_long_colour,
       aes(year, 
           value,
           colour = becoming_more_scientific)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", 
              size = 3,
              colour = "#7570b3") +
  facet_wrap( ~ variable,
              scales = "free_y") +
  theme_bw(base_size = 12) +
  scale_color_manual(values = c("#d95f02", "#1b9e77" )) +
  guides(colour = "none") +
  ylab("")
```

@@

## References

## Acknowledgements

>>>>

To objectively quantify the relative hardness or softness of archaeology, as an evaluation of its status as a science, and the place of this journal in context of other archaeology journals, I take a bibliometric approach. This approach is based on @fanelliBibliometricEvidenceHierarchy2013, who examined the hardness and softness of 12 disciplines using scholarly publication parameters. @fanelliBibliometricEvidenceHierarchy2013 found a spectrum of statistically significant variation in bibliometric variables from the physical to the social sciences, with papers tending at the softer end of the spectrum tending to have fewer co-authors, use less substantive titles, have longer texts, cite older literature, and have a higher diversity of sources. In @fanelliBibliometricEvidenceHierarchy2013's analysis harder sciences include Space Science, Physics, Chemistry, less hard sciences include social sciences include Psychiatry, Psychology, Economics, Business and General Social Sciences, and the Humanities define the soft end of the spectrum. Following @fanelliBibliometricEvidenceHierarchy2013, I quantify the number of authors, length of article, relative title length, age of references, and diversity of references for a large sample of peer-reviewed journal articles.

These parameters are useful because of how they signify consensus in a research community. A larger number of authors on a paper reflects collaboration of people working together on a common goal. Collaborators have specialized roles, each of whom has the ability to study a part of the problem with high accuracy and detail, with harder fields having larger groups of collaborators [@zuckerman1972age]. Reflecting this collaboration group size, harder disciplines tend to have higher average numbers of authors on papers. Article length has an inverse correlation with field hardness. In low-consensus, or softer, fields, papers must be longer to present justification, nuance and contextualization of results. The number of substantive and informative words in an article's title tends to be positively correlated with article length in harder disciplines [@yitzhakiRelationTitleLength2002; @yitzhakiVariationInformativityTitles1997], reflecting a focus on empiricism and efficiency that is characteristic of high-consensus disciplines. While @yitzhakiRelationTitleLength2002 removed stop-words (e.g. prepositions, articles, conjunctions, etc.) to calculate article length, in order to generate results for comparison with @fanelliBibliometricEvidenceHierarchy2013 I follow his method of dividing the total word count of the article title by the total number of pages of the article to compute relative title length.

@@

While Fanelli and Glänzel (2013) analysed papers published in a single year (2012), I found only 303 papers for that same year, and 70.2691554% of papers in the sample published after that date. To make efficient use of the available data and ensure robust representation from different areas of archaeology, including those with lower frequencies of journal article publication, I analysed 9697 papers published during 1975-2025. This sample was collected from Clarivate’s Web of Science database by first selecting the Web of Science category ‘Archaeology’ and the Document type ‘article’ (n = 28,871). To focus on journals of broad relevance to most archaeologists, and that are representative of substantial communities of practice, I then filtered the results to keep only articles published in the top-ranking 25 journals according to their h-indices as reported by Clarivate’s Journal Citation Indicator. Finally I excluded journals with less than 100 articles in the database.

@@

# Number of references ------------------
boxlplot_n_refs <- 
items_df %>% 
  ggplot(aes(1,
             sqrt(refs_n))) +
  geom_boxplot(
               size = 1)  +
  # median for physics, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 6,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 6,
                                label = "p"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  # median for soc sci, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 7,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 7,
                                label = "s"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
    # median for humanities, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 5,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 5,
                                label = "h"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  scale_y_reverse(limits = c(20, 0)) +
  scale_x_continuous(labels = NULL) +
  theme_minimal() +
  theme(panel.grid  = element_blank()) +
  ylab("N. of refs (sqrt)") +
  xlab("Number of references")

# Price's index - age of references ------------------
library(stringr)

@@

plot_grid(boxlplot_n_authors, 
          boxlplot_rel_title_length,
          boxlplot_n_pages,
          boxlplot_n_refs,
          boxlplot_price_index, 
          boxlplot_shannon_index,
          nrow = 2)
```

@fig-mpare-other-fields shows the distribution of bibliometric variables for archaeology in context of data from other fields presented by @fanelliBibliometricEvidenceHierarchy2013. The most striking indicator of archaeology as a hard science is the number of authors, where it is between the social sciences and physics. Archaeology is a close fit with the social sciences in relative title length. It is between the the social sciences and humanities in recency of references and diversity of references. The clearest indicator of archaeology as a soft science is article length where it is similar to the humanities. Overall archaeology does not sit squarely at either end of the hard-soft spectrum. It is generally not a harder science than the social sciences, with the exception of collaborator group sizes.

@@

# compute beta estimates so we can colour lines to indicate more or
# less hard

library(broom)

over_time_long_models <- 
  over_time_long %>%  
  group_nest(variable) %>%  
  mutate(model = map(data, ~tidy(lm(value ~ year, data = .)))) %>%  
  unnest(model) %>%  
  filter(term == 'year') %>%  
  mutate(becoming_more_scientific = case_when( 
    variable == "N. of authors (ln)"         & estimate > 0 ~ "TRUE", 
    variable == "N. of pages (ln)"           & estimate < 0 ~ "TRUE", 
    variable == "N. of refs (sqrt)"          & estimate < 0 ~ "TRUE", 
    variable == "Recency of references"      & estimate > 0 ~ "TRUE", 
    variable == "Relative title length (ln)" & estimate > 0 ~ "TRUE", 
    variable == "Diversity of references"    & estimate < 0 ~ "TRUE",
    TRUE ~ "FALSE"
  ))

# join with data
over_time_long_colour <- 
  over_time_long %>% 
  left_join(over_time_long_models)

```

```{r}
#| label: fig-change-over-time
#| fig-cap: Distribution of article characteristics for archaeology articles over time. D ata points represent individual articles. The Ordinary Least Squares method was used to fit the straight lines summarising the relationships between the variables and the time series.

library(ggpmisc)
formula <-  y ~ x

ggplot(over_time_long_colour, 
       aes(year,  
           value, 
           colour = becoming_more_scientific)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm",  
              size = 3, 
              colour = "#7570b3") +
  stat_poly_eq(use_label("R2", "R2.CI", "P"), 
               formula = formula) +
  facet_wrap( ~ variable,
              scales = "free_y") + 
  theme_bw(base_size = 12) +
  scale_color_manual(values = c("#d95f02", "#1b9e77" )) +
  guides(colour = "none") +
  ylab("") +
  stat_poly_eq(use_label("R2", "R2.CI", "P"), 
               formula = formula)

```

@@
